---
title: "My Project"
author: "Mabvuto Yesaya"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Project description

The data is from TA Wasambo of Karonga district (Northen Malawi). This data is for boreholes/handpumps that people mostly rely on for drinking water collection, secondary to wells and rivers. The data were collected in 2013, and my interests are:

**(placeholder objectives)**

1.  To know the distribution of boreholes per Village.

2.  To check if there is a correlation between those who set tariff fees and the amount that is agreed to be paid as a tariff

3.  To check the different borehole repair parts that are most associated with functionality/non-functionality of boreholes

## Goal

The goal of this document is to explore, clean and select the final variables from borehole Malawi dataset. The selected variables will be used in:

1.   Analysis

2.   Codebook

3.  Making R package

# Import data

```{r}
# library
library(tidyverse)
```

```{r}
## data
#read_csv or read_csv2 to read in a ; separated file does not work for me after consulting perplexity ai for multiple times. In the meantime, I will use read_delim. 

borehole_malawi <- read_delim("/cloud/project/data/raw/Borehole_Malawi_-_all_versions_-_labels_-_2023-12-31-03-34-25.csv",
                            delim = ";")

```
