---
title: "Determinants of borehole' functionality in Northen Malawi"
author:
  - name: "Mabvuto Yesaya"
    orcid: "0000-0003-3234-9699"
    affiliation: "Malawi University of Business and Applied Sciences, Malawi"
    authornote: "ORCID: 0000-0003-3234-9699"
date: "2024-02-10"
bibliography: "../references.bib"
toc: true
output:
  html_document:
    fig_caption: true
    tbl_caption: true
warning: false
code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false

# library
#data wrangling, tidying,column mutations etc
library(tidyverse)

#clean column names
library (janitor)

#for EDA
library(explore)

#here here package
library(here)

#making simple tables
library(knitr)

library(dplyr)

#for making dummy variables
library(fastDummies)

library(ggplot2)
library(unhcrthemes)
library(dplyr)
library(tidyr)
library(scales)

library(treemapify)
library(stargazer)

library(treemap)

library(gtsummary)
library(gt)
```

```{r}
#| echo: false
#load data

borehole_data <- read_csv(here::here ("data/processed", "borehole_malawi_small.csv"))
```

# Introduction

As is the case with least developed countries in the global south, Malawi relies on hand-pumped borehole water for drinking and serves more than 86% of the population [@kumwenda2023triggering; @grimasonClassificationQualityGroundwater2013a]. Studies have found that a significant number of boreholes in Malawi contain elevated levels of contaminants such as fluoride, nitrate, iron, and chlorides, which may exceed the safe limits for drinking water set by the World Health Organization [@chimphamba2014borehole]

A study on handpump borehole functionality was done in TA Wasambo in Karonga district (Northern Malawi) in 2023.  Although the donor community keeps funding these sources of water, @ward2020largescale noted that there are issues with the sustainability of the projects. Therefore, the capstone project aims to explore factors that affect borehole sustainability.  

# Methods

The data were collected in Karonga district, TA Wasambo. Two (2) trained enumerators administered 20-minute questionnaires installed on koboTooolbox ([www.kobotoolbox.org](www.kobotoolbox.org)) to collect borehole functionality data from 300 respondents aged 18 and above. Questions included the borehole maintenance contributions from the community, the availability of borehole spare parts, and borehole management.

The study employed a stratified random sampling technique where 29 villages were purposively sampled based on the availability of boreholes. Then, each borehole water point committee was selected and included in the sample, i.e. one committee could be responsible for multiple villages.

## **Objectives**

1.  To know the users of boreholes per village.

2.  To check if there is a relationship between the costs that were considered before setting the tariff and the amount that is agreed to be paid as a tariff per month now

3.  To check if the actual tariff collected per month, tariff collection frequency, distance to where borehole spare parts are, presence of a service provider, actual tariff collected and conducting preventive maintenance lead to the functionality of boreholes

# Results

## Survey Respondents

A total of 198 surveys were conducted, where each survey targeted a particular borehole. In the villages, as the interest was in the boreholes, the required information was, therefore, sourced from respondents who were better able to supply the required information. The distribution of the survey respondents has been shown in @fig-respondents

```{r}
#| label: fig-respondents
#| fig-cap: "Distribution of respondents"


borehole_data %>%
  group_by(role) %>%
   summarise(n = n()) |> 
  mutate(percent = scales::percent(n/sum(n), accuracy = 1.0)) %>%
  ggplot(aes(area = n, fill = role, label = paste(role, "\n", "(", percent, ")", sep = ""))) +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_text(colour = "black", place = "centre", size = 12,
                    aes(label = paste(role, "\n", "(", percent, ")", sep = ""))) +
  labs(title="Distribution of the survey respondents",
    fill = "Waste categories") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette = "Set3")
```

## Summary statistics

Each borehole has an average of 10 active committee members who look after it, with the boreholes being used by 14-1591 people. The huge variation can be because the boreholes are haphazardly located, and water is drawn from boreholes based on the shortest walking distance. The huge number of people using a borehole means that regular maintenance will be required. The breakdown of the summary statistics is shown in @tbl-stat

```{r}
#| label: tbl-stat
#| tbl-cap: "Summary statistics"

summary <- borehole_data |> 
  select(active_ppo,bh_use_num,tariff_amount) |> 
  na.omit()
table <- summary |> 
  pivot_longer(everything(), names_to = "variable", values_to = "value") 

table|> 
    select(variable,value) |> 
  tbl_summary(by=variable,
              type = list(value ~ "continuous2"),
statistic = all_continuous() ~ c("{mean}", "{sd}", "{min}, {max}", "{median}"),
    missing = "no"
  ) |> 
  modify_header(label = '**Characteristic**',
stat_1 = '**Active comittee members**, N = 89',
stat_2 = '**Households per borehole**, N = 89',
stat_3= '**Tariff (MWK)**, N=89'

) |>  remove_row_type(value, type = "header") |> 
  as_gt()
```

## Distribution of boles per water point committee.

A total of 19 Water Point Committees manage boreholes in TA Wasambo. Each Water Point Committee manages particular boreholes. The distribution of the villagers who draw water managed by a particular water point committee has been shown in @fig-bh

```{r}
#| label: fig-bh
#| fig-cap: "Distribution of boreholes per Water Point Comitee"

borehole_user <- borehole_data %>%
  group_by(village_name) %>%
  summarise(bh_use_num = sum(bh_use_num, na.rm = TRUE))

ggplot(borehole_user) +
  geom_col(aes(
    x = bh_use_num,
    y = reorder(village_name, bh_use_num)
  ),
  fill = unhcr_pal(n = 1, "pal_blue"),
  width = 0.8
  ) +
  geom_text(aes(
    x = bh_use_num,
    y = reorder(village_name, bh_use_num),
    label = round(bh_use_num)
  ),
  hjust = -0.5,
  size = 8 / .pt
  ) +
  labs(
    title = "Households using borehole",
    caption = "Boreholes distributed per a borehole comitee name"
    
  ) +
  scale_x_continuous(expand = expansion(c(0, 0.1))) +
  theme_unhcr(
    grid = FALSE,
    axis = "y",
    axis_title = FALSE,
    axis_text = "y"
  )
```

## Basis for setting a user fee

There were 2 main reasons for setting a user fee. The user fee is set based on what the borehole users can afford or based on the mantainance needs for the system. However, in what ever the basis could be, there is not a difference in the amount collected as the median value is around MWK 500 as shown in @fig-amount

```{r}
#| label: fig-amount
#| fig-cap: "Reasons for setting a tariff"

scatter <- borehole_data %>%
  filter(tariff_amount!=0 & total_money!=0) |> 
  select(tariff_costs_consider,tariff_amount)

ggplot(scatter,aes(x=tariff_costs_consider, y=tariff_amount,fill=tariff_costs_consider))+
  geom_violin()+
  theme(legend.position = "None")+
  labs(
    title="Distribution of the collected user tariff fee",
    x="Basis for setting a tariff system",
    y="Amount of tariff per month (MWK)"
  )+
   theme(plot.title = element_text(size = 16),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13)) 
```

## Functionality of boreholes

Several factors were seen to be corelated with the functionality of boreholes. When the distance to buy spare parts was 20 km away, a 1 unit increase in the distance decreased the probability of borehole functionality by 50 percentage points (p=0.01)0, And if there was no collection of borehole tariff money, the percentage of a borehole being functional also decreased by 50 percentage points. The Model has been shown in @tbl-reg

```{r}
#| label: tbl-reg
#| tbl-cap: Factors associated with borehole functionality.

ols <- borehole_data |> 
  select(available_service_provider:tarrif_frequency_when_required_for_repairs)


mod=lm(bh_functional~ .,data = ols)

stargazer(mod,type = "text",report=("vc*p"),model.names = TRUE)

```

# Conclusions

Many factors come into play as far as the functionality of boreholes is concerned. The more people use a borehole, the more likely it is that the borehole will become non-functional. Also, the fees collected from the boreholes should be regularly collected as agreed, mostly every month. A gap of more than a month in fee collection contributes to borehole non functionality. It is recommended that social enterprises work on bringing borehole spare parts closer to the village (or perhaps right in the village) because when the borehole spare parts are sourced from far away, the chances of non-repair are high.
